---
title: "Module 1: Lab Starter File KEY"
format: html
editor: visual
---

::: callout-important
For this module's lab (and homework), I would like to see your code in the final document. So, make sure you haven't hidden any of your code (i.e., by using `echo=FALSE`) in your code chunks.
:::

## Load Packages

```{r load-packages, message= FALSE}

# type your code here

library(tidyverse)

```

## Read in the Data; View the Data

```{r read-and-view, message= FALSE}

# type your code here

world <- read_csv("data/world_happiness_2015.csv")

View(world)

```

## Exercise 1:

```{r individual-descriptives}

# type your code here

mean_happiness <- mean_happiness <- mean(world$Happiness, na.rm = TRUE)
median_happiness <- median(world$Happiness, na.rm = TRUE)
var_happiness <- var(world$Happiness, na.rm = TRUE)
sd_happiness <- sd(world$Happiness, na.rm = TRUE)

# sanity check: is the sd equal to the sqrt of the variance

sqrt(var(world$Happiness, na.rm = TRUE))
# or
sqrt(var(world$Happiness, na.rm = TRUE)) == sd(world$Happiness, na.rm = TRUE)

```

## Exercise 2:

```{r happiness-summary}

# type your code here

# multiple ways to do it, this is what I like to do; you can always get fancier!
world %>% 
  summarise(mean = mean(Happiness, na.rm = TRUE),
            median = median(Happiness, na.rm = TRUE),
            variance = var(Happiness, na.rm = TRUE),
            standard_deviation = sd(Happiness, na.rm = TRUE))

```

## Exercise 3:

```{r histogram}

# type your code here

world %>% 
  ggplot(aes(x = Happiness)) +
  geom_histogram() +
  labs(title = "World Happiness Ratings in 2015")

# does the variable look normally distributed? skewed? what do the descriptives suggest?

```

## Exercise 4:

```{r skew-kurtosis}

# type your code here

#install.packages("moments")
library(moments)

skew_happiness <- world %>% 
  select(Happiness) %>% 
  skewness(na.rm=TRUE)

kurtosis_happiness <- world %>% 
  select(Happiness) %>% 
  kurtosis(na.rm=TRUE)

```

## Exercise 5:

In the 2015 World Happiness Report, the average Happiness rating is `r round(mean_happiness, 2)`. The median Happiness rating is `r round(median_happiness, 2)`, the variance in Happiness ratings is `r round(var_happiness, 2)`, and the standard deviation is `r round(sd_happiness, 2)`. A histogram of the data reveals an approximately normally distribution. While the skewness of `r round(skew_happiness, 2)` is close to zero and suggests an approximately normal distribution, the kurtosis is slightly high at `r round(kurtosis_happiness, 2)`. For these data, the mean would be the most appropriate measure of central tendency.
