---
title: "Module 1: Homework Starter File"
format: html
editor: visual
---

IMPORTANT: For this module's homework, I would like to see your code in the final document. So, make sure you **HAVE NOT** hidden any of your code (i.e., by using `echo=FALSE`) in your code chunks.

## Load Packages

```{r load-packages, message= FALSE}

# type your code here

library(tidyverse)


```

## Read in and View the Data

```{r read-and-view, message= FALSE}

# type your code here

world <- read_csv("data/world_happiness_2015.csv")

```

## Exercise 1:

Calculate (individually) the mean, median, variance and standard deviation of the "Corruption" variable using the corresponding function in R. Please use these functions:

`mean()` `median()` `var()` `sd()`

Save your values as objects, as we'll be using them in our write-up!

```{r corruption-descriptives}

# type your code here

mean_corrup <- mean(world$Corruption, na.rm = TRUE)
median_corrup <- median(world$Corruption, na.rm = TRUE)
var_corrup <- var(world$Corruption, na.rm = TRUE)
sd_corrup <- sd(world$Corruption, na.rm = TRUE)

```

## Exercise 2:

Construct a **summary** of the "Corruption" variable that contains each of the above parameters.

```{r corruption-summary}

# type your code here

world %>% 
  summarise(mean = mean(Corruption, na.rm = TRUE),
            median = median(Corruption, na.rm = TRUE),
            variance = var(Corruption, na.rm = TRUE),
            standard_deviation = sd(Corruption, na.rm = TRUE))
  


```

## Exercise 3:

Create a histogram of the "Corruption" variable. Make sure to label your plot.

Add a vertical line the indicates the location of the mean (using a solid red line) and median (using a dashed red line). Hint: use `geom_vline`.

What do the descriptive statistics and histogram suggest about the shape of the "Corruption" distribution? Is it normally distributed or skewed in one direction? How can you tell?

```{r histogram}

# type your code here

world %>% 
  ggplot(aes(x = Corruption)) +
  geom_histogram() +
  geom_vline(aes(xintercept = mean_corrup), color = "red") +
  geom_vline(aes(xintercept = median_corrup), linetype = "dashed", color = "red") +
  labs(title = "World Corruption Ratings in 2015")


```

## Exercise 4:

Now, let's examine the skew and kurtosis of the distribution!

First, we need to install and load the `moments` package.

Next, we can use the `skewness()` and `kurtosis()` functions to calculate these values.

Again, save your values as objects, as we'll be using them in the write-up!

```{r skew-kurtosis}

# type your code here

library(moments)

skew_corrup <- world %>% 
  select(Corruption) %>% 
  skewness(na.rm=TRUE)

kurtosis_corrup <- world %>% 
  select(Corruption) %>% 
  kurtosis(na.rm=TRUE)

```

## Exercise 5:

Write up a summary of what you found. Make sure to include information about which measure of central tendency is most appropriate given the distribution. Use the objects you created earlier to insert the descriptive values directly into your summary.

# type your summary here

In the 2015 World Happiness Report, the average Corruption rating is `r round(mean_corrup, 2)`. The median Corruption rating is `r round(median_corrup, 2)`, the variance in Corruption ratings is `r round(var_corrup, 2)`, and the standard deviation is `r round(sd_corrup, 2)`. A histogram of the data reveals an approximately normally distribution. While the skewness of `r round(skew_corrup, 2)` is close to zero and suggests an approximately normal distribution, the kurtosis is slightly high at `r round(kurtosis_corrup, 2)`. For these data, the mean would be the most appropriate measure of central tendency.

## Exercise 6:

\`\`\`{r tidytuesday-descriptives}

# type your code here
