---
title: "Module 3: Homework Starter File"
format: html
editor: visual
---

## Homework Note

You must answer ALL questions and follow ALL steps in the homework instructions!

## Load Packages

```{r load-packages, echo=FALSE, message=FALSE}
library(tidyverse)
library(psych) #for correlation analysis
library(papaja) #for formatting p values
library(broom) #for formatting tables
library(knitr) #for formatting tables
```

## Step 4: Read in the Data

In the first set of exercises, we will use a dataset that looks at the relation between age and pitch perception (in Hz; the ability to distinguish between sounds based on their pitch). You will build a linear model and answer this question: Does pitch perception change as we age?

Read in the data using the `read_csv()` function. Name the data "age_pitch"

```{r read in data, echo = FALSE, message=FALSE}



```

# Part 1

## Exercise 1 (Correlation):

Make a scatter plot of the relation between `age` (x-axis) and `pitch` (y-axis).

Run a correlation test between age and pitch using the `corr.test()` function.

Answer the following questions:

-   Is the correlation positive or negative?
-   Is the correlation strong or weak?

Provide an APA-style summary of your results.

```{r correlation}



```

## Exercise 2 (Regression):

Create a new plot, adding a regression line to your scatter plot of the relationship between `age` (x-axis) and `pitch` (y-axis).

Perform a linear regression analysis using the `lm()` function. Recall the form of regression equations is: `model <- lm(y ~ x, data = df)`.

Answer the following questions:

-   What is the intercept? What does it mean for the relation between age and pitch?
-   What is the slope? What does it mean for the relation between age and pitch?
-   What is the coefficient of determination (R-squared)? What does it mean for the relation between age and pitch?

Create a table of your regression results.

Provide an APA-style summary of your results (make sure it follow the example summary from the lab).

```{r regression}


```

## Exercise 3 (Regression Diagnostics):

Create a histogram of the residuals from your regression (make sure you are plotting the **residuals**, see the lab for information about how to pull these values out of the model).

Next, create a scatter plot with the residuals on the X axis and the predicted values on the Y axis (see homework instructions for code!).

Finally, create a Q-Q plot of the **residuals** using the `qqline()` function in R.

Answer the following questions:

-   Does it appear that the linearity assumption is met? How do you know?
-   Does it appear that the normality assumption is met? How do you know?
-   Does it appear that the constant variance assumption is met? How do you know?

Note: If you detect violations you do not need to do anything to fix them.

```{r diagnostics}



```

# Part 2

Choose a Tidy Tuesday dataset (see Homwork Instructions for details).

## Exercise 4 (Correlation):

Make a scatter plot of the relation between the two variables you chose.

Run a correlation test between age and pitch using the `corr.test()` function.

Answer the following questions:

-   Is the correlation positive or negative?
-   Is the correlation strong or weak?

Provide an APA-style summary of your results.

```{r correlation-2}



```

## Exercise 5 (Regression):

Create a new plot, adding a regression line to your scatter plot of the relation between your two variables.

Perform a linear regression analysis using the `lm()` function. Recall the form of regression equations is: `model <- lm(y ~ x, data = df)`.

Answer the following questions:

-   What is the intercept? What does it mean for the relation between your variables?
-   What is the slope? What does it mean for the relation between your variables?
-   What is the coefficient of determination (R-squared)? What does it mean for the relation between your variables?

Create a table of your regression results.

Provide an APA-style summary of your results (make sure it follow the example summary from the lab).

```{r regression-2}



```

## Exercise 6 (Regression Diagnostics):

First, create a histogram of the residuals from your regression (make sure you are plotting the **residuals**, see the lab for information about how to pull these values out of the model).

Next, create a scatter plot with the residuals on the X axis and the predicted values on the Y axis.

Finally, create a Q-Q plot of the **residuals** using the `qqline()` function in R.

Answer the following questions:

-   Does it appear that the linearity assumption is met? How do you know?
-   Does it appear that the normality assumption is met? How do you know?
-   Does it appear that the constant variance assumption is met? How do you know?

Note: If you detect violations you do not need to do anything to fix them.

```{r diagnostics-2}



```

## Render your document and submit it to Canvas. See Homework Instructions if you need further detail.
