---
title: "Module 4: Homework Starter File"
format: html
editor: visual
---

## Homework Note

You must answer ALL questions and follow ALL steps in the homework instructions!

## Load Packages

```{r load-packages, echo=FALSE, message=FALSE}
library(tidyverse)
library(papaja) #for formatting p values
library(broom) #for formatting tables
library(knitr) #for formatting tables
library(ppcor) #for calculating partial and semi-partial correlations
library(easystats) #for assessing multicollinaerity
```

## Read in the Data

The dataset we will use comes from data from [Kay (2021)](https://www.sciencedirect.com/science/article/pii/S0092656621000659). This paper examines why people believe in conspiracy theories.

The data is called `data_kay2021.csv`. Read in the data using the `read_csv()` function. Name the data "kay2021"

```{r read in data, echo = FALSE, message=FALSE}



```

## Exercise 1:

Say a researcher is interested in whether there is a relationship between the tendency to believe in conspiracy theories (Y = `conspiracy`) and paranoia (X1 = `paranoia`) when controlling for person’s general tendency to have delusions (X2 = `delusions`).

1.  Calculate **and report** the semi-partial and partial correlations for this scenario. For this exercise, your report can be as simple as: "The semi-partial correlation is \_\_\_ and the partial correlation is \_\_\_."

2.  Explain the difference in what the semi-partial versus the partial correlation means.

```{r correlations}




```

## Exercise 2:

We want to predict the tendency to believe in conspiracy theories (Y = `conspiracy`) from paranoia (X1 = `paranoia`) and delusional ideation (X2 = `delusions`). Take the following steps:

1.  Run a regression predicting the tendency to believe in conspiracy theories (Y = `conspiracy`) from paranoia (X1 = `paranoia`) and delusional ideation (X2 = `delusions`). Call it `model_1`.

2.  Write out your interpretations of the overall model (both R<sup>2</sup> and the F-statistic) and each of the regression coefficients ($\beta_0$, $\beta_1$, and $\beta_2$).

3.  Run regression diagnostics (i.e., `check_collinaerity()`) to look for potential multicollinaerity. Comment on what you find, but not need to try to correct anything.

4.  Provide an APA-style summary of your results (make sure it contains **all** necessary pieces, check lab for an example).

```{r regression-1}




```

## Exercise 3:

Treat the model you ran in Exercise 2 as Model 1. Make a new model, Model 2 (name it `model_2`), that includes all the predictors in Model 1 plus the desire for uniqueness (`uniqueness`) and the desire for control (`control`).

Perform a model comparison to test whether the variation in the tendency to believe in conspiracy theories is accounted for by a desire for uniqueness and control (together) *above and beyond* the variation already accounted for by paranoia and delusional ideation.

1.  Run a regression predicting the tendency to believe in conspiracy theories (Y = `conspiracy`) from paranoia (X1 = `paranoia`), delusional ideation (X2 = `delusions`), desire for uniqueness (`uniqueness`) and the desire for control (`control`). Call it `model_2`.

2.  Write out your interpretations of the overall model (both R<sup>2</sup> and the F-statistic) and each of the regression coefficients ($\beta_0$, $\beta_1$, and $\beta_2$).

3.  Run regression diagnostics (i.e., `check_collinaerity()`) on Model 2 to look for potential multicollinaerity. Comment on what you find, but not need to try to correct anything.

4.  Conduct a model comparison between Model 1 and Model 2. Answer the question: Does Model 2 explain significantly more variance in the tendency to believe in conspiracy theories than Model 1?

5.  Report the change in SSE between Model 1 and Model 2. Hint: we called this `ssr` in lab.

6.  Provide an APA-style summary of your results (make sure it contains **all** necessary pieces, check lab for an example).

```{r regression-2}




```

## Render and submit your document.

## **Make sure that you I can see all of your answers in the rendered document!**
