---
title: "Module 2: Lab Starter File"
format: html
editor: visual
---

For this module's lab (and homework), I would like to see your code in the final document. So, make sure you haven't hidden any of your code (i.e., by using `echo=FALSE`) in your code chunks.

## Load Packages

Note: you may have to install these packages first.

```{r load-packages, message= FALSE}

library(tidyverse)
library(infer)
library(knitr)

```

## Read in the Data; View the Data

```{r read-and-view, message= FALSE}

gapminder_2015 <- read_csv("data/gapminder_2015.csv")

```

## Exercise 1:

For this first exercise, we are going to generate 10000 random samples of different sizes (5, 30, and 100) from a skewed distribution. For each sample, we will calculate the mean. Then, we will plot those 10000 means to examine the resulting sampling distributions. We will use the (depressing, sorry!) infant_mortality variable in the gapminder data.

First, let's plot the distribution and generate some descriptives. Create a table of the descriptives (rounded to 2 decimal places).

```{r plot-distribution}

gapminder_2015 %>% 
  ggplot(aes(x = XXX)) +
  XXX() 

mortality_summary <- gapminder_2015 %>% 
  summarise(Mean = XXX,
            Median = XXX, 
            Variance = XXX,
            `Standard Deviation` = XXX)

kable(t(XXX), digits = 2, col.names = c("Statistic", "Value"), caption = "Infant Mortality Descriptive Statistics") 

```

Describe the shape of the distribution. If it is skewed, please specify the direction (positive/right skewed or negative/left skewed).

Next, let's take 1000 samples of size 5 and calculate the mean for each sample. We will use the `rep_sample_n()` function from the *infer* package.

```{r sample-5}

?rep_sample_n

sample_5 <- rep_sample_n(XXX, size = XXX, reps = XXX) %>% #get 100 samples of size 5
  group_by(XXX) %>% #group by sample (what is the "sample" indicator in our data?)
  summarise(mean_infant_mortality = XXX) %>% #now, get the mean for each sample
  mutate(n = XXX) #add a variable that indicates the sample size

```

Now, do the same thing for sample size of 30 and 100 (i.e., copy the above code and change the `size` and `n` to the appropriate values).

```{r sample-30-100}

sample_30


sample_100


```

Next, combine your three data frames together and generate a plot to compare them.

```{r comparison-plot}

all_samples <- bind_rows(sample_5, sample_30, sample_100)

ggplot(XXX) + #use all_samples data
  geom_histogram(aes(x = XXX), color = "black")  + #create a histogram of infant mortality values, make the bars black
  xlab("Infant mortality per 1,000 live births") + #label the x axis
  ylab("Number of countries") + #label the y axis
  XXX(~XXX, ncol = 1) #create a separate graph for each "n" value, put them all in one column



```

Compare the three plots, what does this show?

This illustrates that the distribution approaches normal as sample size increases. Note that, even for a sample size of 5, the *sampling distributon of the mean* is more normal than the original distribution. By the time you get to a sample of 30, it's already looking *much* closer to normal.

## Exercise 2:

Now, let's practice the steps of Null Hypothesis Significance Testing (NHST). There is no data to read in for this exercise.

A psychologist is investigating the impact of a new mindfulness meditation technique on stress levels. It is known that the average stress score on a well-validated scale is 75 for the general population.

The psychologist recruits 25 participants to undergo the mindfulness meditation program for 8 weeks. After the program, the participants complete the standardized stress scale. The sample of participants has an average stress score of 68 with a standard deviation of 12.

**Step 1:** State the null and alternative hypotheses. Use *non-directional* hypotheses.

H~0~: $\mu$ = 75

H~1~: $\mu$ $\neq$ 75

**Step 2:** Choose your alpha level; provide a justification.

$\alpha$: .05

This is the standard alpha level used in Psychology.

**Step 3:** Collect your data.

We're going to pretend we already did this!

Restating from above: Our sample mean is 68 and sample standard deviation is 12. The population mean is 75.

$\bar{x}$ = 68

*s* = 12

$\mu$ = 75

**Step 4:** Define the mean and SEM for the sampling distribution. Estimate the population mean and sd from your data, calculate the SEM. Report the SEM.

*SEM* = $\sigma$ / $\sqrt{n}$

```{r define-distribution}

# first, create objects for each of these values (some of which I'll use later)
population_mean <- XXX
sample_mean <- XXX
estimated_sigma <- XXX
sample_size <- XXX

# then, I'll use those objects to calculate the SEM

SEM <- XXX / sqrt(XXX)


```

**Step 5:** Determine the probability of your data or more extreme under the null distribution. Calculate the t-statistic and then generate a p-value using the `pt()` function.

*t* = ($\bar{x}$ - $\mu$) / *SEM*

```{r t-and-p}

t_statistic <- (XXX - XXX) / XXX

df_value <- XXX - 1

p_value <- 2*pt(abs(XXX), df = XXX, lower.tail = FALSE)

```

**Step 6:** Compare your probability (p-value) to your alpha level and decide whether to reject or retain the null hypothesis.

**Step 7:** Write up your results summary. Make sure to include a Confidence Interval around your estimate of the mean. See Lab Handout for additional notes about what needs to be included in a summary.

95% CI = $\bar{x}$ ±(1.96 x SEM)

```{r confidence-interval}

# remember that you've already created objects for xbar (sample_mean) and the SEM, round to 2 decimal places

# if we didn't know the "critical value" (i.e., 1.96, the value that marks the cutoffs for 95% of the data), here's what we could do:
critical_value <- qnorm(0.975) 

# we use 0.975 instead of 0.95 because we have a two-tailed distribution and want 2.5% of data in each tail (and 95% in the middle between the tails)

# remember that you've already created objects for xbar (sample_mean) and the SEM, round to 2 decimal places

lower_bound <- round(XXX - (1.96*XXX), 2)
upper_bound <- round(XXX + (1.96*XXX), 2)

# OR

lower_bound <- round(XXX - (critical_value*XXX), 2)
upper_bound <- round(XXX + (critical_value*XXX), 2)


```

TYPE SUMMARY HERE

## Consider the following alternative scenarios. What would you conclude?

If observed sample mean was 72, the t statistic would be -1.25 and p-value would be 0.22. What do you conclude?

TYPE CONCLUSION HERE

If the observed sample mean was 82, the t statistic would be 2.92 and p-value would be .008. What do you conclude?

TYPE CONCLUSION HERE

What if p was .06? What would you conclude?

TYPE CONCLUSION HERE
