---
title: "Module 2: Lab Starter File KEY"
format: html
editor: visual
---

For this module's lab (and homework), I would like to see your code in the final document. So, make sure you haven't hidden any of your code (i.e., by using `echo=FALSE`) in your code chunks.

## Load Packages

Note: you may have to install these packages first.

```{r load-packages, message= FALSE}

library(tidyverse)
library(infer)
library(knitr)

```

## Read in the Data; View the Data

```{r read-and-view, message= FALSE}




```

## Exercise 1:

For this first exercise, we are going to generate 10000 random samples of different sizes (5, 30, and 100) from a skewed distribution. For each sample, we will calculate the mean. Then, we will plot those 1000 means to examine the resulting sampling distributions. We will use the (depressing, sorry!) infant_mortality variable in the gapminder data. 


First, let's plot the distribution and generate some descriptives. Create a table of the descriptives (rounded to 2 decimal places). 

```{r plot-distribution}




```
Describe the shape of the distribution. If it is skewed, please specify the direction (positive/right skewed or negative/left skewed).

The data are positively skewed (or right skewed). 

Next, let's take 1000 samples of size 5 and calcualte the mean for each sample. We will use the `rep_sample_n()` function from the *infer* package.

```{r sample-5}




```
Now, do the same thing for sample size of 30 and 100 (i.e., copy the above code and change the `size` and `n` to the appropriate values).

```{r sample-30-100}



```
Next, combine your three data frames together and generate a plot to compare them.

```{r comparison-plot}




```
Compare the three plots, what does this show?

This illustrates that the distribution approaches normal as sample size increases. Note that, even for a sample size of 5, the _sampling distributon of the mean_ is more normal than the original distribution. By the time you get to a sample of 30, it's already looking _much_ closer to normal.


## Exercise 2:

Now, let's practice the steps of Null Hypothesis Significance Testing (NHST). There is no data to read in for this exercise. 

A psychologist is investigating the impact of a new mindfulness meditation technique on stress levels. It is known that the average stress score on a well-validated scale is 75 for the general population.

The psychologist recruits 25 participants to undergo the mindfulness meditation program for 8 weeks. After the program, the participants complete the standardized stress scale. The sample of participants has an average stress score of 68 with a standard deviation of 12.

**Step 1:** State the null and alternative hypotheses. Use _non-directional_ hypotheses.

H~0~: $\mu$ = 75

H~1~: $\mu$ $\neq$ 75

**Step 2:** Choose your alpha level; provide a justification.

$\alpha$: .05

This is the standard alpha level used in Psychology.

**Step 3:** Collect your data.

We're going to pretend we already did this! 

Restating from above: Our sample mean is 68 and sample standard deviation is 12. The population mean is 75. 

$\bar{x}$ = 68

_s_ = 12

$\mu$ = 75

**Step 4:** Define the mean and SEM for the sampling distribution. Estimate the population mean and sd from your data, calculate the SEM. Report the SEM.

_SEM_ = $\sigma$ / $\sqrt{n}$

```{r define-distribution}





```



**Step 5:** Determine the probability of your data or more extreme under the null distribution. Calculate the t-statistic and then generate a p-value using the `pt()` function.

_t_ = ($\bar{x}$ - $\mu$) / _SEM_

```{r t-and-p}





```

**Step 6:** Compare your probability (p-value) to your alpha level and decide whether to reject or retain the null hypothesis. 





**Step 7:** Write up your results summary. Make sure to include a Confidence Interval around your estimate of the mean. See Lab Handout for additional notes about what needs to be included in a summary.

95% CI = $\bar{x}$ &pm;(1.96 x SEM)

```{r confidence-interval}

# remember that you've already created objects for xbar (sample_mean) and the SEM, round to 2 decimal places




```

[type summary here]

## Consider the following alternative scenarios. What would you conclude?

If observed sample mean was 72, the t statistic would be -1.25 and p-value would be 0.22. What do you conclude?

[type conclusion here]

If the observed sample mean was 82, the t statistic would be 2.92 and p-value would be .008. What do you conclude?

[type conclusion here]





