---
title: "Module 9: Lab Starter KEY"
format: html
editor: visual
---

# Purpose

Today's lab will cover categorical x categorical models (i.e, "Factorial ANOVA"). We will go over how to specify interaction terms in R, how to interpret the model output, and how to visualize the results.

For today's lab, you will need to load the following libraries.

```{r libraries, message=FALSE}

library(tidyverse)
library(afex) #for running ANOVA with Type III sums of squares
library(ggrain) #for creating raincloud plots
library(emmeans) #for examining simple effects
library(effectsize) #for calculating effect size
library(broom) #for cleaning up models
library(kableExtra) #for viewing tables produced by `kable` in html
library(Superpower) #for power analysis

```

# Research scenario

Today's dataset examines the effectiveness of two drugs (Drug_A and Drug_B) at three different doses (Low, Medium, and High) in decreasing patients' pain level.

## Read in the data

Today's dataset is called `drug_dose.csv`. Let's read it in and name it `pain`.

```{r read-and-view, message= FALSE, echo=FALSE}

pain <- read_csv("data/drug_dose.csv")

```

## Examine the Data

-   Use `View()` to examine the full dataset
-   Use `str()` to look at the structure of the data
-   Use `head()` to look at the first few rows of the data

```{r message= FALSE, echo=FALSE, eval=FALSE}

View(pain)
str(pain)
head(pain)

```

> Question: What do you notice about the structure of the data? What changes might you want to make?

ANSWER: drug and dose are not factors and we might want to make them factors

# Turn `drug` and `dose` into factors.

Examine the structure of the data to confirm that these changes were made.

```{r message= FALSE, echo=FALSE, eval=FALSE}

pain <- pain %>% 
  mutate(drug = as.factor(drug),
         dose = as.factor(dose))

str(pain)

```

## 2x2 Factorial ANOVA

For our first analysis, let's only compare the `Low` and `Medium` dose levels.

Create a new data object called `pain_red` (pain reduced) in which you have filtered out the `High` dose levels.

```{r message= FALSE, echo=FALSE, eval=FALSE}

pain_red <- pain %>% 
  filter(dose != "High") %>% 
  droplevels() #to remove R's memory of the "High" level 

```

# Let's examine participants' average pain at each level of drug and dose.

Group the data by `drug` and `dose` and get the means.

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: fill in this code to get the means











pain_red %>%
  group_by(drug, dose) %>%
  summarize(pain_level = mean(pain_level)) %>%
  pivot_wider(names_from = "drug",values_from = "pain_level" ) %>%  #Note, here I'm pivoting the table to make it easier to read.
  kable() %>% 
  kable_styling()

```

> Question: From the means, what can you say about the relation between drug and dose in predicting pain?

ANSWER: it looks like maybe there is no effect of drug at Low doses but pain may be higher for Drug_A than Drug_B at Medium doses.

## Linear modeling approach: treatment/dummy-coding

Let's see if this is true, run a regression analysis predicting `pain_level` from `drug`, `dose`, and their interaction.

$$\hat{pain_i} = \beta_0 + \beta_1drug + \beta_2dose + \beta_3drugxdose $$

Run this model using `lm()`.

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: create and run the model








model1 <- lm(pain_level ~ drug * dose, data = pain_red)
summary(model1)

# here we can `tidy` the model and include the confidence intervals

lm(pain_level ~ drug * dose, data = pain_red) %>%
  tidy(conf.int=TRUE) %>% 
  kable() %>% 
  kable_styling()

```

> Question: Write the full model with parameter estimates filled in.

$$\hat{pain_i} = 8.50 - .28(drug) - .79(dose) - .54(drugxdose) $$

> Question: What do each of the parameter estimates mean?

ANSWER:

Plug in different values for combinations into the model.

Intercept: predicted outcome when drug = 0 (Drug_A) and dose = 0 (Low)

Drug: the mean difference in the intercept for a one-unit change in drug when dose = 0; change from the intercept for a low dose for Drug_B

Dose: the mean difference in the intercept for a one-unit change in dose when drug = 0; change from the intercept for Drug_A for a medium dose

Interaction: Does the difference between Drug_A and Drug_B differ across Low and Medium doses?

## Linear modeling approach: deviation-coding

Now, let's deviation code our predictors and run the model again!

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: fill in values for deviation coding










contrasts(pain_red$drug) <- c(0.5, -0.5) # deviation code drug
contrasts(pain_red$dose) <- c(0.5, -0.5) # deviation code dose

#check out the structure of the data to check that this coding took effect
str(pain_red)

model2 <- lm(pain_level ~ drug * dose, data = pain_red)
summary(model2)

# again, we can `tidy` the model and include the confidence intervals

lm(pain_level ~ drug * dose, data = pain_red) %>%
  tidy(conf.int=TRUE) %>% 
  kable() %>% 
  kable_styling()

```

> Question: Write the full model with parameter estimates filled in.

$$\hat{pain_i} = 7.83 - .55(drug) + 1.05(dose) - .54(drugxdose) $$

> Question: What do each of the parameter estimates mean?

ANSWER:

Intercept: the grand mean (overall mean of all the cells)

Drug: the main effect of drug

Dose: the main effect of dose

Interaction: interaction betwen drug and dose

No significant effects.

## ANOVA approach

Now, we will run the same model using the ANOVA approach. We are going to use the `aov_ez` function in the `{afex}` package so that we can use Type III sums of squares!

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: try to fill in the code below










aov_ez(id="subid", between=c("drug", "dose"), dv="pain_level", data=pain_red,  anova_table = list(es = "pes")) %>% 
  nice()%>% 
  kable() %>% 
  kable_styling()

```

> Question: How do the results from the ANOVA compare to the Regression?

ANSWER: They are essentially the same (t\^2 = F)

## 2x3 Factorial ANOVA

Now, we will use the full dataset to compare all levels of dose. First, lets check that the levels of our `dose` variable make sense. If not, change the order.

```{r message= FALSE, echo=FALSE, eval=FALSE}

levels(pain$dose)

#students: change the order of the levels









pain <- pain %>% 
  mutate(dose = factor(dose,
                          levels= c("Low", "Medium", "High")))

```

> Question: What are the three effects that we are testing? What is the null and alternative hypothesis for each test?

ANSWER:

Drug: H0: pain level is equal across all drugs; H1: at least one drug differs Dose: H0: pain level is equal across dosages; H1: at least one dosage differs DrugxDose: H0: the effect of dose does not depend on drug (and vice versa); H1: the effect of dose depends on drug (or vice versa)

Next, run a linear model predicting pain_level from `drug`, `dose`, and their interaction. Make sure you're using the `pain` dataset.

```{r message= FALSE, echo=FALSE, eval=FALSE}

model3 <- lm(pain_level ~ drug * dose, data = pain)
summary(model3)

```

The output becomes a little more challenging to interpret and doesn't give us information about overall effects (e.g., Is there a main effect of drug? ...an interaction between drug and dose?)

One way we can make it more interpretable is using an ANOVA approach. Let's use the `aov_ez` function in the `{afex}` package again to run the ANOVA.

```{r message= FALSE, echo=FALSE, eval=FALSE}

aov_ez(id="subid", between=c("drug", "dose"), dv="pain_level", data=pain,  anova_table = list(es = "pes")) %>% 
  nice()%>% 
  kable() %>% 
  kable_styling()

```

> Question: How would you interpret the effects (remember these are omnibus tests)?

ANSWER: No effect of drug. Difference somewhere in dose. Interaction between drug and dose.

# Let's compare this to the same ANOVA using the `aov()` function in base R.

```{r message= FALSE, echo=FALSE, eval=FALSE}

model_aov <- aov(pain_level ~ drug * dose, data = pain)

summary(model_aov) 

```

> Question: Compare the output of the two ANOVA models. Do they differ? If so, why?

ANSWER: Yes! In particular, `drug` goes from non-significant (with Type III SS) to significant (with Type I SS). This is because Type I asks how much of the *total variance* `drug` explains and Type III asks how much of the variance left over after `dose` and `drug x dose` is explained by `drug`.

## Explore the significant interaction

We found both a significant effect of `dose` and a significant interaction. Because the interaction is significant, we will focus on that (remember that we should use caution in interpreting main effects in the presence of an interaction).

First, let's plot our interaction.

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: fill in the code below to create the plot










ggplot(pain, aes(x = dose, y = pain_level, fill = dose)) +
  geom_rain(rain.side="l") + 
  facet_wrap(~drug) +
  labs(x="Dose", y="Pain Level", title="Effect of Dose by Drug") + #always label your axes!
  theme(legend.position = "none") #we don't need a legend because the X axis is clearly labeled

```

> Question: What does this plot suggest?

ANSWER: For Drug A, high dose levels are really effective at reducing pain while Low and Medium levels don't differ as much. For Drug B there doesn't seem to be a huge difference in pain across the three dosage levels.

Now, we will use emmeans to explore the effect of `dose` at each level of `drug`.

```{r message= FALSE, echo=FALSE, eval=FALSE}

lm(pain_level ~ drug * dose, data = pain) %>%
  emmeans::emmeans(pairwise~drug|dose) %>%
  joint_tests(by="dose") %>%
  kable() %>% 
  kable_styling()

```

Don't forget effect sizes!

```{r message= FALSE, echo=FALSE, eval=FALSE}

#students: plug in the f values for each test below










F_to_eta2(f = c(0.094, 0.781, 15.638), 
          df=c(1,1,1), 
             df_error=45)%>% # for full model
  kable() %>% 
  kable_styling()

```

In our model, the main effect of `dose` was significant. While we are being careful in interpretation due to the significant interaction, maybe we are still curious which levels of dose differ significantly from one another.

We can get this information using emmeans as well!

```{r message= FALSE, echo=FALSE, eval=FALSE}

emmeans::emmeans(model3, ~dose, adjust="bonferroni") %>% 
  test() %>% 
  kable() %>% 
  kable_styling()

```

> Question: Overall, what do these results reveal about the interaction bewteen drug and dosage?

ANSWER: For Drug A, high dose levels are really effective at reducing pain while Low and Medium levels don't differ as much. For Drug B there doesn't seem to be a huge difference in pain across the three dosage levels.

> Question: What information would you include in a summary?

ANSWER: What you did, significant effects (include means, sds, F, df, p-values, effect size, confidence intervals, etc), report non-significant effects but no need to go into detail, general interpretation of what you found. Include a figure depicting the interaction effect.

## ANOVA power

Let's say we want to replicate this study. How would our power be if we had approximately the same sample size?

```{r message= FALSE, echo=FALSE, eval=FALSE}

#first, get means for each condition
pain %>% 
  group_by(drug, dose) %>% 
  summarise(mean = mean(pain_level, na.rm = TRUE),
            sd = sd(pain_level, na.rm = TRUE),
            n = n())

string <- "2b*3b"
n <- 10
# about 10 per condition
mu <- c(8.50, 7.71, 2.50, 8.22, 6.90, 6.14)
# Enter means in the order that matches the labels below.
sd <-c(1.31, 1.80, 1.27, 1.09, 2.08, 3.24)
# SDs
labelnames <- c("drug", "Drug_A", "Drug_B", "dose", "Low", "Medium", "High")
# the label names should be in the order of the means specified above.

#create design object
design_result <- ANOVA_design(design = string,
                   n = n,
                   mu = mu,
                   sd = sd,
                   labelnames = labelnames)

#run simulation
#nsims = 100 is not enough, but we'll do this for now to reduce computing time

simulation_result <- ANOVA_power(design_result, 
                                 alpha_level = .05, 
                                 nsims = 100,
                                 verbose = FALSE)

simulation_result

```

> Question: How much power do we have to detect each of our effects? Is this enough?

ANSWER: Drug: 33%, Dose: 100%, Interaction: 87%

Now, let's look at how much power we would have with different sample sizes (this will take a while to run!)

```{r message= FALSE, echo=FALSE, eval=FALSE}

plot_power(design_result, min_n = 10, max_n = 100)

```

> Question: How many TOTAL participants would we need to achieve 90% power for all of our research questions?

ANSWER: We get 37, but remember that it's 37 per cell of the design. We have 6 cells (each combo of drug and dose), so the total sample size would be 37x6 = 222!
